---
title: "Changing model flexibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Changing model flexibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how a GAM model changes with increased flexibility.
The data we will use here contains tract profiles from diffusion MRI
measurements in a group of patients with Amyotrophic Lateral Sclerosis (ALS)
and a group of matched controls [@Sarica2017].

We start by loading the `tractable` library and `gratia`, as well as `ggplot2`,
which we will use to interpret the fitted model.

```{r setup}
library(tractable)
library(gratia)
library(ggplot2)
```

Next, we will use a function that is included in `tractable` to read this dataset
directly into memory. Importantly, both the group ("ALS" or "CTRL") and the
subject identifier ("subjectID") need to be factors for subsequent analysis
to work properly.

```{r}
sarica <- tractable::read_afq_sarica()
sarica$group <- factor(sarica$class)
sarica$subjectID <- unclass(factor(sarica$subjectID))
```

We fit models with different levels of flexibility, encoded in different values
of `k`:

```{r}
k_values <- c(4, 8, 16, 32)

models <- list()
for (i in 1:length(k_values)){
   models[[i]] <- tractable_single_tract(
    df_afq = sarica,
    tract = "Right Corticospinal",
    participant_id = "subjectID",
    group_by = "group",
    covariates = c("age", "group"),
    dwi_metric = "fa",
    k = k_values[i]
  )
}
```

And we plot the smooths, divided by group for each one of these levels:

```{r}
plots <- list()
for (i in 1:length(k_values)){
  plots[[i]] <- models[[i]] %>%
    gratia::smooth_estimates() %>%
    gratia::add_confint() %>%
    ggplot2::ggplot(aes(x = nodeID, y = .estimate, ymin = .lower_ci,
                        ymax = .upper_ci, group = group, color = group,
                        fill = group)) +
    ggplot2::geom_ribbon(color = NA, alpha = 0.2) +
    ggplot2::geom_line() +
    ggplot2::labs(y = "FA", title = sprintf("k = %d", k_values[i]))
}

plots
```

## References
