---
title: "Changing model flexibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Changing model flexibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how a GAM model changes with increased flexibility.

The data we will use here contains tract profiles from diffusion MRI measurements in a group of patients with Amyotrophic Lateral Sclerosis (ALS) and a group of matched controls [@Sarica2017].

We start by loading the `tractable` library and `gratia`, as well as `ggplot2`, which we will use to interpret the fitted model.

```{r setup}
library(tractable)
library(gratia)
library(ggplot2)
```

Next, we will use a function that is included in `tractable` to read this dataset directly into memory. 

Importantly, both the group ("ALS" or "CTRL") and the
subject identifier ("subjectID") need to be factors for subsequent analysis
to work properly.

```{r}
df_sarica <- read_afq_sarica(na_omit = TRUE) 
df_sarica
```

We fit models with different levels of flexibility, encoded in different values of `k`:

```{r}
k_values <- c(4, 8, 16, 32)

models <- list()
for (i in 1:length(k_values)){
   models[[i]] <- tractable_single_tract(
    target     = "fa", 
    df         = df_cst, 
    tract      = "Right Corticospinal",
    regressors = c("age", "group"), 
    node_k     = k_values[i], 
    node_group = "group"
  )
}
```

And we plot the smooths, divided by group for each one of these levels:

```{r}
plots <- list()
for (i in 1:length(k_values)){
  plots[[i]] <- models[[i]] %>%
    smooth_estimates() %>%
    add_confint() %>%
    filter(.type != "Random effect") %>%
    ggplot(aes(x = nodeID, y = .estimate, ymin = .lower_ci, 
               ymax = .upper_ci, group = group, color = group, 
               fill = group)) +
    geom_ribbon(color = NA, alpha = 0.35) + 
    geom_line(linewidth = 1) +
    scale_y_continuous(name = "FA") +
    ggtitle(sprintf("k = %d", k_values[1])) + 
    theme_bw()
}

plots
```

## References
