---
title: "Including Autocorrelation Effects in GAMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Including Autocorrelation Effects in GAMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the use of an AR1 model within a GAM to account for
autocorrelation of the error terms of the model. We will use data from the
Sarica dataset [@Sarica2017] to demonstrate the how to include an autocorrelation
term and its impact on the model statistics. 

We start by loading the `tractable` library:

```{r setup}
library(tractable)
library(itsadug)
```

Next, we will use a function that is included in `tractable` to read this dataset directly into memory. 

Importantly, both the group ("ALS" or "CTRL") and the
subject identifier ("subjectID") need to be factors for subsequent analysis
to work properly.

```{r load data}
df_sarica <- read_afq_sarica()
df_sarica
```

We will first fit a GAM model without any autocorrelation structure using the 
`tractable_single_bundle` function. This model will use "group" and "age" to 
predict FA, while smoothing over the tract nodes. We will also use the automated
procedure implemented in  `tractable_single_bundle` to determine the ideal value
for `k`, a parameter used to determine the number of spline functions. The default behavior for `tractable_single_bundle` is to include the AR1 model. To avoid this, we set the parameter `autocor` to `FALSE`. 


```{r fit no ac model}
gam_fit_cst_no_ac <- tractable::tractable_single_bundle(
  df_afq         = sarica,
  tract          = "Right Corticospinal",
  participant_id = "subjectID",
  group_by       = "group",
  covariates     = c("age", "group"),
  dwi_metric     = "fa",
  k              = "auto",
  autocor        = FALSE
)
```

Examining the summary of the resulting GAM fit object shows us that the `k=16`
is sufficiently large to describe the spatial variation of tract profile data.
In addition, we see that there is a statistically significant effect of group
(with a p-value of 4.66e-10) and no statistically significant effect of age
(p=0.2748).

```{r get no ac summary}
summary(gam_fit_cst_no_ac)
```

To account for potential spatial autocorrelation of FA values along the length 
of the tract profile, we can incorporate a AR1 model into our GAM. Briefly, this
AR1 model is a linear model that estimates the amount of influence of the FA value of the preceding node on the FA value of the current node (see [@VanRij2019](https://journals.sagepub.com/doi/pdf/10.1177/2331216519832483) for an overview of accounting for autocorrelation using the `mgcv` package). 

The AR1 model takes a parameter $\rho$ to estimate autocorrelation effects. We 
can pass our initial model into the function `itsadug::start_value_rho` to 
automatically determine the value of $\rho$. We can also plot the autocorrelation
by setting `plot=TRUE` within that function (pictured below). 

```{r calculate rho}
rho = itsadug::start_value_rho(gam_fit_cst_no_ac)
itsadug::start_value_rho(gam_fit_cst_no_ac, plot=T)
```

By default, the `tractable_single_bundle` function will determine the value of 
$\rho$ and incorporate the AR1 model into the GAM estimation. 

```{r fit model including AR structure}
gam_fit_cst <- tractable::tractable_single_bundle(
  df_afq         = sarica,
  tract          = "Right Corticospinal",
  participant_id = "subjectID",
  group_by       = "group",
  covariates     = c("age", "group"),
  dwi_metric     = "fa",
  k              = "auto"
)
```

Examining the summary of the resulting GAM fit object shows us that the inclusion
of the AR1 model changes the resulting statistics of our model. Although there 
is still a statistically significant effect of group (p=0.044), the value of the
t-statistic on this term has changed from 6.243 to 2.015. 

```{r summarise AR model}
summary(gam_fit_cst)
```

## References
