---
title: "Including Autocorrelation Effects in GAMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Including Autocorrelation Effects in GAMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the use of an AR1 model within a GAM to account for
autocorrelation of the error terms of the model. We will use data from the
Sarica dataset [@Sarica2017] to demonstrate the how to include an autocorrelation term and its impact on the model statistics. 

We start by loading the `tractable` library:

```{r setup}
library(tractable)
library(itsadug)
```

Next, we will use a function that is included in `tractable` to read this dataset directly into memory. 

Importantly, both the group ("ALS" or "CTRL") and the subject identifier ("subjectID") need to be factors for subsequent analysis to work properly.

```{r load_data}
df_sarica <- read_afq_sarica(na_omit = TRUE)
df_sarica
```

We will first fit a GAM model without any autocorrelation structure using the 
`tractable_single_tract` function. This model will use "age" and "group" to 
predict FA, while smoothing over the tract nodes. We will also use the automated
procedure implemented in  `tractable_single_tract` to determine the ideal value
for `k`, a parameter used to determine the number of spline functions. The default behavior for `tractable_single_tract` is to include the AR1 model. To avoid this, we set the parameter `autocor` to `FALSE`. 


```{r fit_no_autocor_model}
cst_fit_no_autocor <- tractable_single_tract(
  df         = df_sarica,
  tract      = "Right Corticospinal",
  target     = "fa",
  regressors = c("age", "group"),
  node_group = "group",
  k          = "auto",
  autocor    = FALSE
)

cst_no_autocor_summary <- summary(cst_fit_no_autocor)
cst_no_autocor_summary
```

Examining the summary of the resulting GAM fit object shows us that the `k=16`
is sufficiently large to describe the spatial variation of tract profile data.
In addition, we see that there is a statistically significant effect of group
(with a p-value of `r cst_no_autocor_summary$p.table["groupCTRL", "Pr(>|t|)"]`) and no statistically significant effect of age
(p = `r cst_no_autocor_summary$p.table["age", "Pr(>|t|)"]`).

To account for potential spatial autocorrelation of FA values along the length 
of the tract profile, we can incorporate a AR1 model into our GAM. Briefly, this
AR1 model is a linear model that estimates the amount of influence of the FA value of the preceding node on the FA value of the current node (see [@VanRij2019](https://journals.sagepub.com/doi/pdf/10.1177/2331216519832483) for an overview of accounting for autocorrelation using the `mgcv` package). 

The AR1 model takes a parameter $\rho$ to estimate autocorrelation effects. We 
can pass our initial model into the function `itsadug::start_value_rho` to 
automatically determine the value of $\rho$. We can also plot the autocorrelation by setting `plot = TRUE` within that function (pictured below). 

```{r calculate_rho}
rho <- start_value_rho(cst_fit_no_autocor)
start_value_rho(cst_fit_no_autocor, plot = TRUE)
```

By default, the `tractable_single_tract` function will determine the value of 
$\rho$ and incorporate the AR1 model into the GAM estimation. 

```{r fit_autocor_model}
cst_fit <- tractable_single_tract(
  df         = df_sarica,
  tract      = "Right Corticospinal",
  target     = "fa",
  regressors = c("age", "group"),
  node_group = "group",
  k          = "auto"
)

cst_summary <- summary(cst_fit)
cst_summary
```

Examining the summary of the resulting GAM fit object shows us that the inclusion of the AR1 model changes the resulting statistics of our model. Although there  is still a statistically significant effect of group (p = `r cst_summary$p.table["groupCTRL", "Pr(>|t|)"]`), the value of the t-statistic on this term has changed from `r cst_no_autocor_summary$p.table["groupCTRL", "t value"]` to `r cst_summary$p.table["groupCTRL", "t value"]`. 

You can also examine the AR1 model residuals for autocorrelations with the `itsadug::acf_resid` function. 

```{r plot_corrected_acf}
acf_resid(cst_fit)
```


## References
