---
title: "Fitting Generalized Additive Models (GAMs)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting Generalized Additive Models (GAMs)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the use of GAMs for statistical analysis of
tract profile data. The data we will use here contains tract profiles from
diffusion MRI measurements in a group of patients with Amyotrophic Lateral
Sclerosis (ALS) and a group of matched controls [@Sarica2017].

We start by loading the `tractable` library:

```{r setup}
library(tractable)
```

Next, we will use a function that is included in `tractable` to read this dataset directly into memory. 

Importantly, both the group ("ALS" or "CTRL") and the
subject identifier ("subjectID") need to be factors for subsequent analysis
to work properly.

```{r}
df_sarica <- read_afq_sarica(na_omit = TRUE)
df_sarica
```

First, let's visualize the data. We use the `plot_tract_profiles` function,
selecting to view both fractional anisotropy (FA) and mean diffusivity profiles
in two tracts: the right corticospinal tract (CST) and the right superior
longitudinal fasciculus (SLF), which are identified in the "tractID" column of
this dataset.

```{r}
plot_tract_profiles(
  df         = df_sarica,
  group_col  = "group",
  n_groups   = 2,
  metrics    = c("fa", "md"),
  tracts     = list("Right Corticospinal", "Right SLF"),
  tracts_col = "tractID"
)
```

We can already see that ALS has a profound effect on the tract profiles of the
CST, but does not affect SLF as much. We will use GAMs in order to quantify
this in statistical terms. We start by fitting a GAM model to the data from the
CST. Using the `tractable_single_bundle` function, we select the Right CST data,
and focus here only on FA. We use "group" and "age" as relevant covariates.
Comparing group as a main effect, that will also be used to fit separate
smooth functions for each category of subjects. The `mgcv` GAM functions
use a parameter `k` to determine how many spline functions to use in fitting
the smooth change of FA over the length of the tract. We use an automated
strategy to find `k`.

```{r}
cst_fit <- tractable_single_tract(
  df         = df_sarica,
  tract      = "Right Corticospinal",
  target     = "fa",
  regressors = c("age", "group"),
  node_group = "group",
  k          = "auto"
)
```

Examining the summary of the resulting GAM fit object shows us that the `k=16`
is sufficiently large to describe the spatial variation of tract profile data.
In addition, we see that there is a statistically significant effect of group
(with a p-value of 4.66e-10) and no statistically significant effect of age
(p=0.2748).

```{r}
summary(cst_fit)
```

Running the same analysis on the data from SLF, we see that there is no
significant difference between the groups in this bundle, indicating that the
effect observed in CST is rather specific to this bundle.

```{r}
slf_fit <- tractable_single_tract(
  df         = df_sarica,
  tract      = "Right SLF",
  target     = "fa",
  regressors = c("age", "group"),
  node_group = "group",
  k          = "auto"
)

summary(slf_fit)
```

This is in line with other work that we have conducted with this dataset
using other methods [@Richie-Halford2021SGL].

If you are interested in exploring the ALS dataset even more, you can also
see this data in an AFQ-Browser [here](yeatmanlab.github.io/Sarica_2017/)
[@Yeatman2018AFQB].

## References
